### 2019-07-30 paper.tex
Initial import

\documentclass[10pt]{article}
%\documentclass[10pt,twocolumn]{article}
\usepackage[a4paper]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{titling}
\usepackage{graphicx}

\title{\textbf{Towards File System integrity verification in BTRFS}}
\author{
	\textbf{Johannes Thumshirn}\\
	\normalsize SUSE Labs - Filesystems Team\\
	\normalsize \href{mailto:jthumshirn@suse.de}{jthumshirn@suse.de}
}
\date{}

\begin{document}
\maketitle

\begin{abstract}
	\textbf{
	Integrity verification of a computer's File System has recently become
	more important as it is not always possible to ensure a File System
	has not been altered between two mounts. This is especially important
	if the administrator has no control over the physical computer anymore
	like in a cloud environment or with embedded systems and the Internet
	of things.
	\\In this paper I present a solution for use with the BTRFS File System.}
\end{abstract}
\section{Introduction} 
With Linux being more and more present in the emerging cloud and embedded
systems markets extra precaution has to be taken that an administrator or
service owner can ensure the File System she uses has not been altered by a
malicious third party.

In contrast to a normal enterprise server or laptop environment, the
administrator does not always have full control over the File System and thus
can not take an unaltered File System for guaranteed.  Even for mobile
computing environments there is the so called "Evil maid attack" \cite{maid},
which is the attack of an unattended device by a physical attacker, with the
goal to alter the device in a way later gain access to the data stored on this
device or the device itself.

This can either be done by altering the internal structure of the File System,
for instance by exchanging the
inode\footnote{\href{https://en.wikipedia.org/wiki/Inode}
{https://en.wikipedia.org/wiki/Inode}} number of the \textbf{/bin/login} login
service to \textbf{/bin/true}, a program which has the sole purpose of
returning the boolean value of \textbf{true}. Another possible version of the
"Evil Maid Attack" targeting disk encryption is to replace the program
decrypting the disk with a version controlled by the attacker, that a)
decrypts the hard disk with the user-supplied key but b) records the key as
well so the attacker can then later extract it and access the data.

For applications of Linux like in embedded systems, with hosting co-location
or virtual machines running in a public cloud environment, there is next to no
possibility for an administrator to control whether a File System has been
altered or not. To check if a single file has been replaced with a malicious
version the \textit{Linux Kernel's Integrity Measurement
Architecture}\cite{IMA} can be utilized. IMA records a cryptographic hash of a
file and saves it as out-of-band information in an extended attribute
\footnote{\href{https://en.wikipedia.org/wiki/Extended\_file\_attributes}
{https://en.wikipedia.org/wiki/Extended\_file\_attributes}} of the file.
Saving cryptographic hash of a file protects the file from being altered,
however it does not protect the File Systems internal structures from being
altered.

%% Add transition to next section

\section{File System Architecture}
BTRFS\cite{BTRFS} is a general purpose File System for the Linux Operating
system that was first introduced in 2009. BTRFS is based on a copy-on-write
\footnote{\href{https://en.wikipedia.org/wiki/Copy-on-write\#In\_computer\_storage}
{https://en.wikipedia.org/wiki/Copy-on-write\#In\_computer\_storage}} capable
B-tree\cite{rodeh} data-structure. The main features of BTRFS are
copy-on-write, File System level snapshots, subvolumes, data and meta-data
checksums, transparent compression and multi-device storage pools. BTRFS is
mainly developed by Facebook, Fujitsu, SUSE and
Oracle\footnote{\href{https://btrfs.wiki.kernel.org/index.php/Contributors}
{https://btrfs.wiki.kernel.org/index.php/Contributors}} but has had
contributions from other companies as well in the past.

In BTRFS updates to a block on disk don't happen to the to-be updated block
directly but the block first is updated in the host's memory and then written
elsewhere on disk. After the block has been written, the File System internal
data structures used to locate this block get updated in the same fashion as
the block itself, meaning the data is first read into the host's memory,
updated and then written to a new location on disk. This way the File System
is always guaranteed to be consistent in an event of a power loss or eventual
disk failure.

Another feature targeting disk failure is the fact that all on-disk structures
of the File System have a checksum which protects the data. Figures
~\ref{fig:btrfs-node} and ~\ref{fig:btrfs-leaf} present the
\texttt{btrfs\_node} and \texttt{btrfs\_leaf} data structures which are the
basic building block for on-disk meta-data in BTRFS. These data structures
starts with a \texttt{btrfs\_header} embedded which is shown in Figure
~\ref{fig:btrfs-header}. The \texttt{btrfs\_header} data structure embedded in
\texttt{btrfs\_node} and \texttt{btrfs\_leaf} begins with a \texttt{checkusm}
field protecting not only the \texttt{btrfs\_header} but the whole
\texttt{btrfs\_node} and \texttt{btrfs\_leaf} data structure, including the
offset and size fields in \texttt{btrfs\_leaf} which tell the position of the
data portion of the searched item in the leaf as well as the blockptr in
\texttt{btrfs\_node}, which holds the address of other blocks in the File
System.

Accompanying this checksum for meta-data, there is a checksum for every data
block on the File System and these checksums are written to disk in a special
area. This area can be accessed by the File System's checksum-tree, which is
like any other tree in the BTRFS File System a B-tree that has the same
copy-on-write update semantics the other meta-data nodes in the File System
have.

The fact that BTRFS already has checksums for both, the data saved in the
File System as well as the meta-data comprising the File System and it's
strict copy-on-write nature, meaning there are only a few occasions where data
on disk is overridden, enables it for being transformed into a File System
which can be authenticated an verified.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{btrfs-node.pdf}
	\caption{BTRFS Node Structure}
	\label{fig:btrfs-node}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{btrfs-leaf.pdf}
	\caption{BTRFS Leaf Structure}
	\label{fig:btrfs-leaf}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{btrfs-header.pdf}
	\caption{BTRFS Header Structure}
	\label{fig:btrfs-header}
\end{figure}

\section{Architecture}
\subsection{Preliminary Work}
After the initial rework of BTRFS' checksum handling was merged in Linux
Kernel version 5.3, a second patchset has been sent to add a second checksum
type XXHASH64\footnote{\href{http://www.xxhash.org}{http://www.xxhash.org}}
that was frequently requested by the community. XXHASH64 is no
cryptographically secure hash function but it can be used to ease
de-duplication of blocks and is generally considered a fast hash function. Table
~\ref{tab:hash-speed} holds performance data of Hash functions evaluated by
David Sterba in the course of this.
This patchset also included the necessary changes to BTRFS' user-space
utilities in order to create a File System with a different checksum algorithm
than the default CRC32C.

\begin{table}
	\centering
	\begin{tabular}{l | r | r } 
		Hash & Total Cycles & Cycles per Iteration \\	
		\hline
		CRC32C & 1783997556 & 1783 \\
		CRC32C\_SW & 1738434898 & 1738 \\
		XXHASH & 2501653700 & 2501 \\
		BLAKE2b-256-arch & 17754335778 & 17754 \\
		BLAKE2b-256 & 23867725764 & 23867 \\
		BLAKE2s-arch & 25877188085 & 25877 \\
		BLAKE2s & 34630837258 & 34630 \\
		SHA256 & 106150926755 & 106150 \\
		SHA3 & 289233228792 & 289233 \\
		\hline
	\end{tabular}
	\caption{Speedtest of different hash functions with 4096 byte blocks
	and 1,000,000 iterations}
	\label{tab:hash-speed}
\end{table}

\subsection{Architectural Considerations}
\subsubsection{Cryptographic Hash}
This variant exploits the fact, that each data block's checksum is saved in
the checksum tree. As this tree is an ordinary btrfs meta-data tree the leaf
nodes save the data block checksums, the intermediate nodes save the block
pointers to these checksums and get checksummed as well up to the tree root,
which then saves the checksum for it's block pointers (which point to the
intermediate nodes or leaf nodes of the checksum tree). \\ Using this scheme
we can save the checksum of the checksum tree's root node on every unmount and
on every mount we can check if these checksums still match. If they do not
match, we could refuse the mount.\\ The checksum root's checksum could for
instance be stored inside a TPM
chip\footnote{\href{https://en.wikipedia.org/wiki/Trusted\_Platform\_Module}
{https://en.wikipedia.org/wiki/Trusted\_Platform\_Module}} and before mounting
it could be retrieved from the TPM.\\ The downside to this is, on every
unmount, the new hash value needs to be stored safely and on every mount the
on-disk hash has to be checked against the known good hash.

\subsubsection{HMAC}
This variant takes variant 1 a step further by using an hash-based message
authentication code \cite{HMAC} instead of a plain hash function. HMACs use an
user-supplied key and mix it into the calculation of the hash. The key used
for this could be safely stored in a TPM or any other secure enclave and would
not need to be modified.\\
The downside of this is the higher computational efforts needed by employing
an HMAC as opposed to a plain cryptographically secure hash function.\\
Another downside is the fact, that the used key has to be deployed at
File System creation time and cannot easily be modified afterwards.\\
This approach is favoured by most involved parties.

\subsubsection{Hybrid Hash with HMAC on Root Node}
This variant is a hybrid approach between the above two proposals. In this
variant, we calculate a cryptographic hash as checksum for every data and
meta-data block and let then calculate the HMAC for the root node of the
checksum tree.\\ This way we save the overhead of the HMAC's double checksum
calculation but still get the advantage of only having to save the key one for
a verification of the File System.

\subsection{File System Verification}
If the ``plain'' HMAC approach will be taken, the next step after the inclusion
of the XXHASH64 patchset will be implementing a new mount option, which
queries the Linux Kernel's internal keyring for a key which will be used to
create the HMACs. Also the respective HMAC function needs to be enabled in
BTRFS and the user-space code needs to be adopted for this as well.

\section{Conclusion}
TODO

\section{Related Work}
This paper does not show a novel approach to the trust verification of a 
File System, but builds upon lessons learned from earlier approaches.

\subsection{dm-verity}
Dm-verity is a Linux device-mapper target calculating a rolling hash on the
block device level. As dm-verity works on a block device level it needs an
extra partition or block device to save the resulting hash tree, or a reserved
area on the same block device can be used. Dm-verity is also designed to be
read-only.\\
when dm-verity is used as between the File System and the physical block
device, for every read of a data block on the device a hash is calculated and
it is verified up to the root node of the hash tree. Only if the whole hash
chain matches, the data is returned. Otherwise an IO error is signaled to the
upper layers.

\subsection{fs-verity}
Fs-verity works similar to dm-verity, but instead of a hash tree for the whole
block device, fs-verity has a Merkle-Tree for every file in the File System
hidden after the last block of the file, but hidden from the normal
File System operations and the user.

\subsection{UBIFS}
TODO

\subsection{ZFS}
TODO

\begin{thebibliography}{9}
\bibitem{BTRFS}
	Ohad Rodeh, Josef Bacik and Chris Mason.
	\textit{BTRFS: The Linux B-tree Filesystem}
	\\\texttt{https://domino.research.ibm.com/library/cyberdig.nsf\\/papers/6E1C5B6A1B6EDD9885257A38006B6130/\$File/rj10501.pdf}
	\\IBM Research, Fusion-IO.
\bibitem{rodeh}
	Ohad Rodeh.
	\textit{B-trees, shadowing, and clones}
	\\\texttt{\href{https://www.usenix.org/legacy/events/lsf07/tech/rodeh.pdf}{https://www.usenix.org/legacy/events/lsf07/tech/rodeh.pdf}}
	USENIX Linux Storage \& Filesystem Workshop 2007. Also Rodeh, Ohad (2008). "B-trees, shadowing, and clones". ACM Transactions on Storage.
\bibitem{maid}
	Wikipedia.
	\textit{Evil Maid Attack}
	\\\texttt{\href{https://en.wikipedia.org/wiki/Evil\_maid\_attack}{https://en.wikipedia.org/wiki/Evil\_maid\_attack}}
\bibitem{IMA}
	Linux IMA Project
	\textit{IMA - Integrity Measurement Atchitecture}
	\\\texttt{\href{http://linux-ima.sourceforge.net/}{http://linux-ima.sourceforge.net/}}
\bibitem{HMAC}
	H. Krawczyk, M. Bellare, R. Canetti
	\textit{HMAC: Keyed-Hashing for Message Authentication}
	\\\texttt{\href{https://tools.ietf.org/html/rfc2104}{https://tools.ietf.org/html/rfc2104}}
	Request for Comments: 2104 	
\end{thebibliography}
\end{document}

### 2019-07-30 paper.tex
Add transition between paragraphs.
Update fs-verity section.

@@
 
-%% Add transition to next section
+In order to protect the File System's internal data structures as well as the
+data stored in the File System both the internal structure and the payload
+have to secured. A File System which has the possibility to do this, is BTRFS.
 
@@
 \subsection{fs-verity}
-Fs-verity works similar to dm-verity, but instead of a hash tree for the whole
-block device, fs-verity has a Merkle-Tree for every file in the File System
-hidden after the last block of the file, but hidden from the normal
-File System operations and the user.
+Fs-verity\cite{FS-VERITY} works similar to dm-verity, but instead of a hash
+tree for the whole block device, fs-verity has a Merkle-Tree for every file in
+the File System hidden after the last block of the file, but hidden from the
+normal File System operations and the user. The two differences to IMA are,
+fs-verity a) uses a Merkle-Tree instead of a single hash, to speed up open(2)
+calls and b) uses free space at the end of a file instead of an extended
+attribute as this Merkle-Tree can grow larger than the maximum size of an
+extended attribute. Fs-verity does not protect the File System's internal data
+structures from being altered.
 
@@
 	Request for Comments: 2104 	
+\bibitem{FS-VERITY}
+	J. Corbet
+	\textit{Protecting files with fs-verity}
+	\\\texttt{\href{https://lwn.net/Articles/763729/}{https://lwn.net/Articles/763729/}}
 \end{thebibliography}

### 2019-07-30 paper.tex
Move Conclusion section and add body.

@@
 
-\section{Conclusion}
-TODO
-
 \section{Related Work}
@@
 
+\section{Conclusion}
+Thanks to BTRFS already using checksums for both data and meta-data, turning
+it into an authenticated and verified File System is a technically feasible
+task.  Using BTRFS instead of other File System or Block Layer level
+mechanisms also provides advantages, like authenticated meta-data, as opposed
+to fs-verity, or the ability to be able to alter the File System by a
+legitimate administrator or user of the File System.
+
 \begin{thebibliography}{9}

### 2019-07-30 paper.tex
Add paragraph for UBIFS.

@@
 \subsection{UBIFS}
-TODO
+UBIFS is the Linux Kernel's File System typically used in the Embedded Device
+Market, where raw flash chips are used as opposed to traditional Hard Disk
+Drives or Solid State Disks.\\
+UBIFS also has the capability of authenticating the File System. Similar to
+what is proposed for BTRFS in this paper, UBIFS is using a HMAC to protect
+it's internal on-disk data structures as well as the file data stored on the
+File System.
 

### 2019-07-30 paper.tex
Clarify fs-verity's upstream and fs support status

@@
 extended attribute. Fs-verity does not protect the File System's internal data
-structures from being altered.
+structures from being altered.\\
+As of today (Linux Kernel v5.3-rc), fs-verity is only implemented for the ext4
+and f2fs File Systems and not merged into the mainline kernel.
 

### 2019-07-30 paper.tex
Add bibliography entries for UBIFS.

@@
 \subsection{UBIFS}
-UBIFS is the Linux Kernel's File System typically used in the Embedded Device
-Market, where raw flash chips are used as opposed to traditional Hard Disk
-Drives or Solid State Disks.\\
-UBIFS also has the capability of authenticating the File System. Similar to
-what is proposed for BTRFS in this paper, UBIFS is using a HMAC to protect
-it's internal on-disk data structures as well as the file data stored on the
-File System.
+UBIFS\cite{UBIFS} is the Linux Kernel's File System typically used in the
+Embedded Device Market, where raw flash chips are used as opposed to
+traditional Hard Disk Drives or Solid State Disks.\\ UBIFS also has the
+capability of authenticating the File System. Similar to what is proposed for
+BTRFS in this paper, UBIFS is using a HMAC to protect it's internal on-disk
+data structures as well as the file data stored on the File System.
 
@@
 	\\\texttt{\href{https://lwn.net/Articles/763729/}{https://lwn.net/Articles/763729/}}
+\bibitem{UBIFS}
+	S. Hauer.
+	\textit{UBIFS Authentication}
+	\\\texttt{\href{https://www.kernel.org/doc/Documentation/filesystems/ubifs-authentication.md}{https://www.kernel.org/doc/Documentation/filesystems/ubifs-authentication.md}}
 \end{thebibliography}

### 2019-07-30 paper.tex
Add dots to bibitems.
Add ZFS section.

@@
 \subsection{ZFS}
-TODO
+ZFS\cite{ZFS} is a File System originally developed at Sun Microsystems (now
+Oracle) for the Solaris Operating System. ZFS uses a Merkle Tree\cite{ZFS} on
+the block level with either the Fletcher2 or SHA-256 Hash algorithm. This
+Merkle Tree is embedded into the File System's internal structure and in a
+way that the File System's Superblock (or Uberblock in ZFS Terms) constantly
+provides an up-to-date signature of the blocks in the File System.
 
@@
 \bibitem{IMA}
-	Linux IMA Project
+	Linux IMA Project.
 	\textit{IMA - Integrity Measurement Atchitecture}
@@
 \bibitem{FS-VERITY}
-	J. Corbet
+	J. Corbet.
 	\textit{Protecting files with fs-verity}
@@
 	\\\texttt{\href{https://www.kernel.org/doc/Documentation/filesystems/ubifs-authentication.md}{https://www.kernel.org/doc/Documentation/filesystems/ubifs-authentication.md}}
+\bibitem{ZFS}
+	Oracle Corporation.
+	\textit{ZFS End-to-End Data Integrity}
+	\\\texttt{\href{https://blogs.oracle.com/bonwick/zfs-end-to-end-data-integrity}{https://blogs.oracle.com/bonwick/zfs-end-to-end-data-integrity}}
 \end{thebibliography}

### 2019-07-30 paper.tex
Reorder companies

@@
 checksums, transparent compression and multi-device storage pools. BTRFS is
-mainly developed by Facebook, Fujitsu, SUSE and
-Oracle\footnote{\href{https://btrfs.wiki.kernel.org/index.php/Contributors}
+mainly developed by SUSE, Facebook, Oracle and Fujitsu
+\footnote{\href{https://btrfs.wiki.kernel.org/index.php/Contributors}
 {https://btrfs.wiki.kernel.org/index.php/Contributors}} but has had

### 2019-08-02 paper.tex
Extent section about the verification algorithm.

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
@@
 \subsection{File System Verification}
-If the ``plain'' HMAC approach will be taken, the next step after the inclusion
-of the XXHASH64 patchset will be implementing a new mount option, which
-queries the Linux Kernel's internal keyring for a key which will be used to
-create the HMACs. Also the respective HMAC function needs to be enabled in
-BTRFS and the user-space code needs to be adopted for this as well.
+If the ``plain'' HMAC approach will be taken, the next step after the
+inclusion of the XXHASH64 patchset will be implementing a new mount option,
+which queries the Linux Kernel's internal keyring for a key which will be used
+to create the HMACs upon mount time. Also the respective HMAC function needs
+to be enabled in BTRFS and the user-space code needs to be adopted for this as
+well.\\
+
+For ever data block $b$ and a given key $K$ an HMAC is calculated according to
+the following formula:
+\[
+HMAC(K,b) = H((K' \oplus opad) || H((K' \oplus ipad)|| b ))
+\]
+\[
+K =
+\begin{cases}
+	H(K) & \text{if } K \text{is larger than the hash blocksize} \\
+	K & \text{otherwise}
+\end{cases}
+\]
+With $opad$ and $ipad$ being the defined outer and inner paddings defined by
+the HMAC RFC. These values will be saved in the File System's Checksum tree
+for each data block. The internal on-disk structure's will use the above
+formula for their checksum values stored in \texttt{struct btrfs\_header}.\\
+
+If a data block has been modified, without the modifier having access to the
+HMAC key $K$ a read of this data block will be rejected by the File System's
+end-io hooks for the Page Cache's \texttt{readpage()} method and an error
+message will be printed to the Kernel's log. The same will happen for disk
+blocks representing File System internal meta-data. Despite it's name the
+\texttt{\_\_readpage\_endio\_check()} function also gets called when the Page
+Cache is avoided, like in the case of Direct-IO and thus the integrity and
+authenticity of the data is guaranteed as well.
 

### 2019-08-02 paper.tex
Remove trailing whitespace.
use graph instead of tabular representation for algorithm performance

@@
 de-duplication of blocks and is generally considered a fast hash function. Table
-~\ref{tab:hash-speed} holds performance data of Hash functions evaluated by
+~\ref{fig:hash-speed} holds performance data of Hash functions evaluated by
 David Sterba in the course of this.
@@
 
-\begin{table}
+\begin{figure}
 	\centering
-	\begin{tabular}{l | r | r } 
-		Hash & Total Cycles & Cycles per Iteration \\	
-		\hline
-		CRC32C & 1783997556 & 1783 \\
-		CRC32C\_SW & 1738434898 & 1738 \\
-		XXHASH & 2501653700 & 2501 \\
-		BLAKE2b-256-arch & 17754335778 & 17754 \\
-		BLAKE2b-256 & 23867725764 & 23867 \\
-		BLAKE2s-arch & 25877188085 & 25877 \\
-		BLAKE2s & 34630837258 & 34630 \\
-		SHA256 & 106150926755 & 106150 \\
-		SHA3 & 289233228792 & 289233 \\
-		\hline
-	\end{tabular}
+	\includegraphics[width=0.9\textwidth]{algo-perf-data.png}
 	\caption{Speedtest of different hash functions with 4096 byte blocks
 	and 1,000,000 iterations}
-	\label{tab:hash-speed}
-\end{table}
+	\label{fig:hash-speed}
+\end{figure}
 
